INGRID: Creating Languages in MPS from ANTLR Grammars

### 2017-06-12 chapters/background.tex
adding my todo notes

\todo{For each MPS term (Structure, Editor, TextGen, concept, and so on) say which generally/widely known terms (from related academic work) are similar/analogous. Relate the MPS-specific terms to the large body of academic work.}

@@

\todo{Mozna pouzit termin "language concept" misto samotneho "concept".}

@@

The building blocks of MPS models are called nodes.
Code of any program in MPS is built from nodes, which represent instances of concepts defined in one of the languages that the program declares to be using.
\todo{zvyraznit slovo "concept"} In the context of MPS, a concept is, in fact, a building block of a language definition.
We use the terms \emph{MPS concept} and \emph{AST node} when needed to avoid confusion.

@@

The definition of an MPS concept (language element) consists of several aspects, where each aspect codifies a different part of the AST nodes' behavior.
The essential aspects are the following: \emph{Structure} (abstract syntax), \emph{Editor} (concrete syntax) and \emph{TextGen}.
Editor defines the concrete syntax (i.e., how the code is visualized and edited) and TextGen specifies how AST nodes are transformed into textual representation.
If, instead of generating text directly, programs in the language are supposed to be transformed into another language that is available in MPS, the Generator aspect must be used to specify the model-to-model transformation rules.
Since only the Structure, Editor and TextGen aspects are relevant for the contribution of this paper, we describe these three below in more detail, and neglect other aspects such as type system and data-flow.

@@

\paragraph{TextGen.}
The TextGen aspect specifies how a given AST node will be translated into plain text representation.
It is typically needed only for the bottom-line base languages.
DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since these are rarely converted to text directly.
After TextGen has generated textual sources from an AST, a compiler for the particular GPL is invoked to compile the textual sources into binary code.
The TextGen definition follows a very straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes at the right moments.
MPS calls the corresponding method on the root AST nodes of the given program.

@@

\subsection{Related Projects}
\label{sect:RELATED}

@@

Besides that, we are aware of three other projects that provide certain support for GPLs in the context of MPS --- namely PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
We describe their main features and limitations in this section.

@@

PE grammars extend the syntax of the ANTLR v4 notation by constructs that provide information about the possible layout of AST nodes.
The current version, as of November 2016, supports just horizontal lists, vertical lists, and some indentation rules.
However, even these few features make the syntax much more complicated.

@@

\todo{
S. Erdweg, T. Rendel, C. K\"{a}stner, and K. Ostermann. Layout-Sensitive Generalized Parsing. In Proceedings of SLE 2012, LNCS 7745.
	Source code layout specified declaratively through constraints on shape and relative positions of syntax trees (AST).
	Layout constraints are represented/encoded using annotations on production rules of context-free grammars.
	Useful for layout-sensitive programming language (e.g., Python).
	Their proposed approach: grammar rule applies only if the input satisfied the constraint (on shape and positioning of source code fragments).
		Parsing algorithm that accepts only inputs with valid layout (that satisfies the constraints attached to grammar rules).
	Constraints support (can express): indentation, number of line breaks, alignment.
	Hand-coded approach used in Python (how the annotated grammar is processed) --- mentioned for comparison.
		Special tokens that indicate layout aspects (indentation, alignment) are added to stream of tokens by preprocessor.
		Layout handled by algorithm that works with syntax trees (ASTs) and token streams.
	The main difference from our approach (INGRID): layout inferred automatically (via automated learning) from several input files.
		No layout annotations in the grammar are used.
		Currently, simple heuristics are used to create editor and textgen.
}

>>>>

\todo{For each MPS term (Structure, Editor, TextGen, concept, and so on) say which generally/widely known terms (from related academic work) are similar/analogous. Relate the MPS-specific terms to the large body of academic work.}
\todo{Mozna rict jak se zakladni aspekty (Structure, Editor, TextGen) nazyvaji v jinych IDE (Spoofax, apod).}

@@

\todo{Mozna pouzit termin "language concept" misto samotneho "concept" - ale jen pri definici.}

@@

The building blocks of MPS models are called nodes.
Code of any program in MPS is built from nodes, which represent instances of concepts from the languages that the program declares to be using.
\todo{zvyraznit slovo "concept"} In the context of MPS, a concept is, in fact, \todo{a language element or} a building block of a language definition.
We use the terms \emph{MPS concept} and \emph{AST node} when needed to avoid confusion.

@@

The definition of an MPS concept (language element) consists of several aspects, where each aspect codifies a different part of the AST nodes' behavior.
The essential aspects are the following: \emph{Structure} (abstract syntax), \emph{Editor} and \emph{TextGen}.
Editor defines the concrete syntax (i.e., how the code is visualized and edited) and TextGen specifies how AST nodes are transformed into textual representation.
If, instead of generating text directly, programs in the language are supposed to be transformed into another language that is available in MPS, the Generator aspect must be used to specify the model-to-model transformation rules.
Since only the Structure, Editor and TextGen aspects are relevant for the contribution of this paper, we describe these three below in more detail, and neglect other aspects such as type system and data-flow.

@@

\paragraph{TextGen.}
The TextGen aspect specifies how a given AST node will be translated into plain text representation.
It is typically needed only for the bottom-line base languages.
DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since \todo{programs in these languages} are rarely converted to text directly.
After TextGen has generated textual sources from an AST, a compiler for the particular GPL is invoked to compile the textual sources into binary code.
The TextGen definition follows a very straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes at the right moments.
MPS calls the corresponding method on the root AST nodes of the given program.

@@

\subsection{Related Work}
\label{sect:RELATED}

\todo{We are aware of only few attempts to create ports of general-purpose languages into (IDEs based on) projectional editors/editing. All of them were conducted manually.}

@@

Besides that, we are aware of three other projects that provide certain support for GPLs in the context of MPS --- namely PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
We describe their main features and limitations in this section.
\todo{Check what has been done for Spoofax.}

\todo{Pro vsechny nasledujici projekty (PE4MPS, ANTLR{\_}MPS, mps{\_}metabnf) mention GPPLs for which there projects have been used (e.g., JavaScript). We discuss application of PE4MPS on JavaScript in Section 4.}

@@

PE grammars extend the syntax of the ANTLR v4 notation by constructs that provide information about the possible layout of AST nodes.
\todo{Check if there is a new (more recent) version with more features.} The current version, as of November 2016, supports just horizontal lists, vertical lists, and some indentation rules.
However, even these few features make the syntax much more complicated.

@@

\todo{Najit dalsi podobne systemy (treba Spoofax) a kratce popsat existujici reseni podobnych uloh (automated import of general-purpose languages) v jejich kontextu.}

\todo{
S. Erdweg, T. Rendel, C. K\"{a}stner, and K. Ostermann. Layout-Sensitive Generalized Parsing. In Proceedings of SLE 2012, LNCS 7745.
	Source code layout specified declaratively through constraints on shape and relative positions of syntax trees (AST).
	Layout constraints are represented/encoded using annotations on production rules of context-free grammars.
	Useful for layout-sensitive programming language (e.g., Python).
	Their proposed approach: grammar rule applies only if the input satisfied the constraint (on shape and positioning of source code fragments).
		Parsing algorithm that accepts only inputs with valid layout (that satisfies the constraints attached to grammar rules).
	Constraints support (can express): indentation, number of line breaks, alignment.
	Hand-coded approach used in Python (how the annotated grammar is processed) --- mentioned for comparison.
		Special tokens that indicate layout aspects (indentation, alignment) are added to stream of tokens by preprocessor.
		Layout handled by algorithm that works with syntax trees (ASTs) and token streams.
	The main difference from our approach (INGRID): No layout annotations in the grammar are used. Currently, simple heuristics are used to create editor and textgen (i.e., to determine layout).
}

### 2017-06-12 chapters/introduction.tex
adding my todo notes

\todo{
Napsat ze input ANTLR grammar has to be tweaked before INGRID can process them.
Za tim bude "We evaluated INGRID on complex mainstream languages, such as JavaScript and C\#. Time needed for adjusting of the mainstream languages is short ... (less than one hour).
}

@@

The rest of the paper has the following structure.
In the next section, we provide necessary information about MPS and a brief overview of related work, including projects with similar goals as \textsc{Ingrid}.
Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.
We also discuss our experience with usage of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\#, and highlight open problems (Section~\ref{sect:EVAL}), and then we conclude.

>>>>

\todo{
Napsat ze input ANTLR grammar has to be tweaked before INGRID can process them.
Za tim bude "We evaluated INGRID on complex mainstream languages, such as JavaScript and C\#. Time needed for adjusting of the mainstream languages is short ... (less than one hour).
Rict nejak zkracene nasledujici text/vetu: Therefore, the resulting MPS language typically still has to be adjusted manually to improve its usability, and the human users must also define the remaining aspects not yet supported by \textsc{Ingrid}.
}

@@

The rest of the paper has the following structure.
In the next section, we provide necessary information about MPS and a brief overview of related work, including projects with similar goals as \textsc{Ingrid}.
\todo{Specifically, in Section~\ref{sect:MPS} we provide all information necessary to understand the examples from MPS presented in the paper.}
Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.
We also discuss our experience with usage of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\#, and highlight open problems (Section~\ref{sect:EVAL}), and then we conclude.

### 2017-06-12 chapters/language_definition.tex
new format
adding my todo notes

Our approach differs from the related projects (Section~\ref{sect:RELATED}) especially in the level of automation. It also has much better support for the definition of Editor and TextGen.

@@

The grammar of SimpleXML contains elements of the kinds that are listed below.
Each color in Figure~\ref{fig:SIMPLEXML} corresponds to a different kind in a way that is indicated by the list item headers.
\begin{itemize}
	\item \textbf{ANTLR v4 keywords} are required by the notation.
	\item \antlrparserrule{\textbf{Parser rules}} describe the structure of a language.
		Alternatives on the right side of a rule are separated by the pipe character ($|$).
	\item \antlrlexerrule{\textbf{Lexer rules}} describe terminal symbols that the parser matches against the input string.
		A terminal symbol can be encoded as a string value or using a regular expression.
	\item \antlrliteralnoap{\textbf{String literals}}, always written inside of a pair of single quote marks, also represent terminal symbols, but by exact match to a string constant.
	\item \antlrregex{\textbf{Regular expressions}} describe string tokens to be matched against regular expressions with a special ANTLR v4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
\end{itemize}
Elements on the right side of a rule can be annotated with standard EBNF operators (\code{?}, \code{+}, \code{*}) that specify the allowed number of occurrences.

@@

\begin{figure}[ht]
\begin{alltt}
\small
\mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2}.\textbf{size} > 0) \{
  \mpstgaction{append} \textcolor{Blue}{\textbf{\textbackslash{}n;}}
  \mpstgaction{indent buffer;}
  \mpstgkeyword{with indent} \{
    \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2} with  \};
  \}
  \mpstgaction{append} \textcolor{Blue}{\textbf{\textbackslash{}n;}}
\} 
\end{alltt}
\caption{Fragment of the TextGen aspect of \mpsconcept{Element{\_}1} with adjusted indentation}
\label{fig:TEXTGENADJUSTED}
\end{figure}

>>>>

Our approach differs from the \todo{existing} related projects (Section~\ref{sect:RELATED}) especially in the level of automation. It also has much better support for the definition of Editor and TextGen.

@@

The grammar of SimpleXML contains elements of the kinds that are listed below.
Each color in Figure~\ref{fig:SIMPLEXML} corresponds to a different kind in a way that is indicated by the list item headers.
\begin{itemize}
	\item \textbf{ANTLR v4 keywords} are required by the notation.
	\item \antlrparserrule{\textbf{Parser rules}} describe the structure of a language.
		Alternatives on the right side of a rule are separated by the pipe character ($|$).
	\item \antlrlexerrule{\textbf{Lexer rules}} describe terminal symbols that the parser matches against the input string.
		A terminal symbol can be encoded as a string value or using a regular expression.
	\item \antlrliteralnoap{\textbf{String literals}} \todo{tohle urcite ztmavit}, always written inside of a pair of single quote marks, also represent terminal symbols, but by exact match to a string constant.
	\item \antlrregex{\textbf{Regular expressions}} \todo{tohle mozna ztmavit} describe string tokens to be matched against regular expressions with a special ANTLR v4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
\end{itemize}
Elements on the right side of a rule can be annotated with standard EBNF operators (\code{?}, \code{+}, \code{*}) that specify the allowed number of occurrences.

@@

\begin{figure}[ht]
\begin{alltt}
\small
\mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2}.\textbf{size} > 0) \{
  \mpstgaction{append} \textcolor{blue}{\textbf{\textbackslash{}n;}}
  \mpstgaction{indent buffer;}
  \mpstgkeyword{with indent} \{
    \mpstgaction{append} \$list\{\mpstgparam{node}.\mpstgnodeprop{\textit{Content}}{\_}\textit{2} with  \};
  \}
  \mpstgaction{append} \textcolor{blue}{\textbf{\textbackslash{}n;}}
\} 
\end{alltt}
\caption{Fragment of the TextGen aspect of \mpsconcept{Element{\_}1} with adjusted indentation}
\label{fig:TEXTGENADJUSTED}
\end{figure}
